Can the symmetric \textit{Fermi} and eROSITA bubbles be produced by tilted jets?

### 2022-04-24 paper.tex
update

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[]{Can the symmetric \textit{Fermi} and eROSITA bubbles be produced by tilted jets: a proof-of-concept study}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[P. H. Tseng et al.]{\large
Po-Hsun Tseng$^{1}$%
\thanks{\href{mailto:zengbs@gmail.com}{zengbs@gmail.com}}%
\orcidlink{https://orcid.org/0000-0002-1868-0660}%
\quad H.-Y. Karen Yang$^{2,3,4,5}$%
\thanks{\href{mailto:hyang@phys.nthu.edu.tw}{hyang@phys.nthu.edu.tw}}%
\orcidlink{https://orcid.org/0000-0003-3269-4660}%
\quad Hsi-Yu Schive$^{1,5,6,7}$%
\thanks{\href{mailto:hyschive@phys.ntu.edu.tw}{hyschive@phys.ntu.edu.tw}}%
\orcidlink{https://orcid.org/0000-0002-1249-279X}%
\quad Chun-Yen Chen$^{8}, $
\quad Tzihong Chiueh$^{1,5,6}$%
\thanks{\href{mailto:chiuehth@phys.ntu.edu.tw}{chiuehth@phys.ntu.edu.tw}}%
\orcidlink{https://orcid.org/0000-0003-2654-8763}%
\\\\
$^{1}$Institute of Astrophysics, National Taiwan University, Taipei 10617, Taiwan\\
$^{2}$Institute of Astronomy, National Tsing Hua University, Hsinchu 30013, Taiwan\\
$^{3}$Department of Physics, National Tsing Hua University, Hsinchu 30013, Taiwan\\
$^{4}$Center for Informatics and Computation in Astronomy, National Tsing Hua University, Hsinchu 30013, Taiwan\\
$^{5}$Physics Division, National Center for Theoretical Sciences, Taipei 10617, Taiwan\\
$^{6}$Department of Physics, National Taiwan University, Taipei 10617, Taiwan\\
$^{7}$Center for Theoretical Physics, National Taiwan University, Taipei 10617, Taiwan\\
$^{8}$Institute of physics, National Taiwan University, Taipei 10617, Taiwan\\
}

@@

% Abstract of the paper
\begin{abstract}
The \textit{Fermi Gamma-Ray Space Telescope} reveals two large bubbles in the Galaxy,\
which extend nearly symmetrically $\sim50^{\circ}$ above and below the Galactic center (GC).\
The recent discovery of giant eROSITA bubbles also shows such a symmetry about the GC, suggesting that they may originate from a single GC activity.\
Previous simulations of bubble formation that invoke active galactic nucleus (AGN) jets\
have assumed that the jets are vertical to the Galactic plane;\
however, in general there does not need to be a correlation between the jet orientation and the rotational axis of the Galactic disk.\
Using three-dimensional (3D) special relativistic hydrodynamic (SRHD) simulations that\
include cosmic rays (CRs) and thermal gas, we show that the interaction between the AGN jets and the thin interstellar medium (ISM) disk is crucial for producing the symmetric \textit{Fermi} and eROSITA bubbles (collectively, Galactic bubbles).\
%We also produce simulated gamma-ray and\
%microwave spectra from the inverse Compton (IC) scattering and synchrotron,\
%and compare them with the observed counterparts in order to verify\
%an oblique AGN jets model.
Specifically, we find that\
(1) the simulated Galactic bubbles are nearly symmetric about the GC albeit\
the bipolar jets are at an angle $45^{\circ}$ with respect to\
the rotational axis of the Galaxy;\
(2) the edge of the eROSITA bubbles corresponds to a forward shock,\
originally driven by oblique bipolar jets emanating from the GC 12.39 million years ago; %,\
%and significantly stretched by the stratified atmosphere afterwards.\
(3) followed by the forward shock is a tangling contact discontinuity,\
which we interpret as the edge of the \textit{Fermi} bubbles\
composed of turbulent and high-temperature ($\sim2$ keV)\
plasma in pressure balance with the external medium;
(4) assuming a leptonic model with spatially uniform CR electron spectrum, we find that the observed gamma-ray and microwave spectra can be reproduced with a best-fit CR power-law index of 2.4.\
The broad agreement between simulated and observed multi-wavelength features suggests that forming the Galactic bubbles by oblique AGN jets is a plausible scenario.
%has thus released a caveat:\
%the jets shall be vertical to the disk normal.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
editorials, notices -- miscellaneous
\end{keywords}

@@

\section{Discussion}
Our modeled gamma-ray and microwave spectra assumed that
the underlying CRe spectrum is spatially uniform with hard spectral index 2.4;
however, the high energy CRe severely suffer from synchrotron and IC losses,
during their passage through the magnetic and radiation fields, respectively, within the Galaxy.

The typical synchrotron and IC cooling time scale of high energy ($\sim100$ GeV) CRe
in Milky Way is $\sim$ 1 Myr \citep{Yang2017}, ten times shorter than
12.39 Myr required for the simulated bubbles to reach the desired morphology.

Therefore, the CRe generating the gamma-ray emission would need to be re-accelerated
by in-situ acceleration mechanisms such as shocks or turbulence.
Since the forward shock is rather far away from the gamma-ray bubbles,
re-acceleration is more likely associated with turbulence \citep{Mertsch2011,Mertsch2019},
possibly in balance with the IC and synchrotron cooling. We will investigate the competition between stochastic acceleration and radiative cooling
in future work.

@@

\item In the oblique jet scenario, the edge of the eROSITA bubbles is a forward shock,\
      originally driven by oblique bipolar jets emanating from the GC 12.39 Myr ago for a duration of 1.2 Myr\
      and significantly expanded within the stratified atmosphere afterwards.\
      The overall extent of the simulated X-ray bubbles is comparable to that of the eROSITA bubbles, though not as limb-brightened.\
      Future models including shock-accelerated CRs may help to resolve this issue by increasing the compressibility of the fluid and enhancing the thermal Bremsstrahlung emissivity at the edge of the X-ray bubbles.
      %The simulated eROSITA bubbles are not as limb-brightened as the observation.\
      %A possibility to enhance the X-ray emission is to include shock-accelerated CRs near the shock,\
      %in which CRs could increase the compressibility of the fluid,\
      %resulting in the enhanced thermal Bremsstrahlung emissivity that is proportional to density squared.
\item Followed by the forward shock is a turbulent, hot plasma ($\sim2$ keV) filled with the injected CRs in pressure balance with the shocked medium, which we interpret as the \textit{Fermi} bubbles.\
      The similarity of the morphology, temperature, and gamma-ray spectrum between\
      simulated and observed bubbles suggests that\
      the edge of the \textit{Fermi} bubbles are likely associated with the contact discontinuity between the turbulent, hot plasma and the post-shock medium. The surface of the simulated bubbles is not as smooth as the observed ones; inclusion of magnetic fields in the future may help suppress the instabilities at the bubble surface due to the mangetic draping effect.
\item Assuming a leptonic model with a spatially uniform, power-law CRe spectrum ranging from 0.5 MeV to 562.1 GeV, we showed that the observed gamma-ray and microwave spectra can simultaneously reproduced. The best-fit\
      CRe power-law index is found to be 2.4.
\item The elapsed time of 12.39 Myr required for the simulated bubbles
      to reach the desired morphology
      is ten times longer than the typical synchrotron and IC cooling time scale of
      high energy ($\sim100$ GeV) CRe. Thus, re-acceleration of CRe by shocks or turbulence must be considered
      in this model. Since the forward shock is rather far away from the gamma-ray bubbles, stochastic acceleration of CRs by turbulence appears to be more plausible.\
      We will investigate the competition between stochastic acceleration and radiative cooling
      in a future work.
      %{\color{red} KY: This item is to be modified according to the newly added text in the Discussion section. PH: done!}
\end{itemize}

\section{Acknowledgements}
The authors thank Mateusz Ruszkowski and Ellen G. Zweibel for insightful comments.\
We thank Henry Zovaro for discussions about the innermost bubbles\
during the final stages of this work.\
We thank Peter Predehl for providing the range of observed X-ray intensity in Fig. \ref{fig__xray_0.8keV_angle_000}.\
A part of the simulations are performed and analyzed using computing resources\
operated by the National Center for High-Performance Computing (NCHC).
HYKY acknowledges support from\
Yushan Scholar Program of the Ministry of Education of Taiwan\
and\
Ministry of Science and Technology of Taiwan (MOST 109-2112-M-007-037-MY3).
HS acknowledges\
funding support from the Jade Mountain Young Scholar Award No. NTU-109V0201,\
sponsored by the Ministry of Education, Taiwan.\
This research is partially supported by the Ministry of Science and\
Technology of Taiwan (MOST) under grants MOST 107-2119-M-002-036-MY3\
and MOST 108-2112-M-002-023-MY3, and the NTU\
Core Consortium project under grants NTU-CC-108L893401 and\
NTU-CC-108L893402.

@@

Finally, we find that increasing the source temperature (B$\rightarrow$C$\rightarrow$D)\
not only improves the symmetry of the bubbles but also increases the temperature
inside the bubbles at the present time. This result implies that a high-temperature source
helps the jets ablate the clumpy dense gas away and pours into the atmosphere
through the intercloud channels of the clumpy disk. In contrast,
if source temperature is not high enough,
the jets cannot effectively ablate the high-density clouds within 12.39 Myr,
resulting in the formation of asymmetric bubbles (case B).

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[]{Can the symmetric \textit{Fermi} and eROSITA bubbles be produced by tilted jets?}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[P. H. Tseng et al.]{\large
Po-Hsun Tseng$^{1}$%
\thanks{\href{mailto:zengbs@gmail.com}{zengbs@gmail.com}}%
\orcidlink{https://orcid.org/0000-0002-1868-0660}%
\quad H.-Y. Karen Yang$^{2,3,4,5}$%
\thanks{\href{mailto:hyang@phys.nthu.edu.tw}{hyang@phys.nthu.edu.tw}}%
\orcidlink{https://orcid.org/0000-0003-3269-4660}%
\quad Hsi-Yu Schive$^{1,5,6,7}$%
\thanks{\href{mailto:hyschive@phys.ntu.edu.tw}{hyschive@phys.ntu.edu.tw}}%
\orcidlink{https://orcid.org/0000-0002-1249-279X}%
\quad Tzihong Chiueh$^{1,5,6}$%
\thanks{\href{mailto:chiuehth@phys.ntu.edu.tw}{chiuehth@phys.ntu.edu.tw}}%
\orcidlink{https://orcid.org/0000-0003-2654-8763}%
\\\\
$^{1}$Institute of Astrophysics, National Taiwan University, Taipei 10617, Taiwan\\
$^{2}$Institute of Astronomy, National Tsing Hua University, Hsinchu 30013, Taiwan\\
$^{3}$Department of Physics, National Tsing Hua University, Hsinchu 30013, Taiwan\\
$^{4}$Center for Informatics and Computation in Astronomy, National Tsing Hua University, Hsinchu 30013, Taiwan\\
$^{5}$Physics Division, National Center for Theoretical Sciences, Taipei 10617, Taiwan\\
$^{6}$Department of Physics, National Taiwan University, Taipei 10617, Taiwan\\
$^{7}$Center for Theoretical Physics, National Taiwan University, Taipei 10617, Taiwan\\
}

@@

% Abstract of the paper
\begin{abstract}
The \textit{Fermi Gamma-Ray Space Telescope} reveals two large bubbles in the Galaxy,\
which extend nearly symmetrically $\sim50^{\circ}$ above and below the Galactic center (GC).\
The recent discovery of giant eROSITA bubbles also shows such a symmetry about the GC,
suggesting that they may originate from a single GC activity.\
Previous simulations of bubble formation that invoke active galactic nucleus (AGN) jets\
have assumed that the jets are vertical to the Galactic plane;\
however, in general there does not need to be a correlation between
the jet orientation and the rotational axis of the Galactic disk.\
Using three-dimensional (3D) special relativistic hydrodynamic (SRHD) simulations that\
include cosmic rays (CRs) and thermal gas, we show that {\color{red}the clumpy gas within ISM\
disrupts jets collimation and confinement (``failed jets" hereafter),
which in turn causes the failed jets to form
hot bubbles and subsequent bouyancy makes it vertical to produce
the symmetric \textit{Fermi} and eROSITA bubbles (collectively, Galactic bubbles).}
Specifically, we find that\
(1) the simulated Galactic bubbles are nearly symmetric about the GC albeit\
the bipolar jets are at an angle $45^{\circ}$ with respect to\
the rotational axis of the Galaxy;\
(2) the edge of the eROSITA bubbles corresponds to a forward shock,\
originally driven by oblique bipolar jets emanating from the GC around 12 million years ago; %,\
(3) followed by the forward shock is a tangling contact discontinuity,\
which we interpret as the edge of the \textit{Fermi} bubbles\
composed of turbulent and high-temperature ($\sim2$ keV)\
plasma in pressure balance with the external medium;
(4) assuming a leptonic model with spatially uniform CR electron spectrum,
we find that the observed gamma-ray and microwave spectra can be reproduced with
a best-fit CR power-law index of 2.4. The broad agreement between simulated and
observed multi-wavelength features suggests that forming the Galactic bubbles by
oblique AGN failed jets is a plausible scenario.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
Relativistic hydrodynamics, \textit{Fermi} bubbles, eROSITA bubbles, cosmic rays
\end{keywords}

@@

\section{Discussion}
{\color{red}
As a past violent event at the GC must destroy the Central Molecular Zone (CMZ)
but the CMZ is currently either in a quasi-steady state or
presently increasing \citep{Crocker2012,Krumholz2015,Sormani2019};
we conclude that the time since the violent event must be ``slightly" shorter
than the reconstruction time of CMZ on a timescale of
$M_{\text{CMZ}}/\dot{M}_{\text{inflow}}$, where $M_{\text{CMZ}}$ is
the total mass of molecular gas within $R\sim300$ pc,
and $\dot{M}_{\text{inflow}}$ is the gas inflow rate from the Galactic disc at
$R\sim3$ kpc down to the outskirts of the CMZ.
Since $M_{\text{CMZ}}$ and $\dot{M}_{\text{inflow}}$ have been reported to be in the range
$2-6\times10^{7}$ M$_{\odot}$ \citep{Dahmen1998,Ferrire2007} and
$0.4-2.7$ M$_{\odot}$ yr$^{-1}$ \citep{Sormani2019,Crocker2012}, respectively;
we estimate that the CMZ formation is about a couple of 10 Myrs, similar to
12 Myr required for our simulated bubbles to reach the desired morphology.
}

On the other hand, our modeled gamma-ray and microwave spectra assumed that
the underlying CRe spectrum is spatially uniform with hard spectral index 2.4;
however, the high energy CRe severely suffer from synchrotron and IC losses,
during their passage through the magnetic and radiation fields, respectively, within the Galaxy.

The typical synchrotron and IC cooling time scale of high energy ($\sim100$ GeV) CRe
in Milky Way is $\sim$ 1 Myr \citep{Yang2017}, ten times shorter than
12 Myr suggested by our simulation.
Therefore, the CRe generating the gamma-ray emission would need to be re-accelerated
by in-situ acceleration mechanisms such as shocks or turbulence.
Since the forward shock is rather far away from the gamma-ray bubbles,
re-acceleration is more likely associated with turbulence \citep{Mertsch2011,Mertsch2019},
possibly in balance with the IC and synchrotron cooling. We will investigate the
competition between stochastic acceleration and radiative cooling in future work.

@@

\item In the oblique jet scenario, the edge of the eROSITA bubbles is a forward shock,\
      originally driven by oblique bipolar jets emanating from the GC around 12 Myr ago for a duration of 1.2 Myr\
      and significantly expanded within the stratified atmosphere afterwards.\
      The overall extent of the simulated X-ray bubbles is comparable to that of the eROSITA bubbles, though not as limb-brightened.\
      Future models including shock-accelerated CRs may help to resolve this issue by increasing the compressibility of the fluid and enhancing the thermal Bremsstrahlung emissivity at the edge of the X-ray bubbles.
      %The simulated eROSITA bubbles are not as limb-brightened as the observation.\
      %A possibility to enhance the X-ray emission is to include shock-accelerated CRs near the shock,\
      %in which CRs could increase the compressibility of the fluid,\
      %resulting in the enhanced thermal Bremsstrahlung emissivity that is proportional to density squared.
\item Followed by the forward shock is a turbulent, hot plasma ($\sim2$ keV) filled with the injected CRs in pressure balance with the shocked medium, which we interpret as the \textit{Fermi} bubbles.\
      The similarity of the morphology, temperature, and gamma-ray spectrum between\
      simulated and observed bubbles suggests that\
      the edge of the \textit{Fermi} bubbles are likely associated with the contact discontinuity between the turbulent, hot plasma and the post-shock medium. The surface of the simulated bubbles is not as smooth as the observed ones; inclusion of magnetic fields in the future may help suppress the instabilities at the bubble surface due to the mangetic draping effect.
\item Assuming a leptonic model with a spatially uniform, power-law CRe spectrum ranging from 0.5 MeV to 562.1 GeV, we showed that the observed gamma-ray and microwave spectra can simultaneously reproduced. The best-fit\
      CRe power-law index is found to be 2.4.
\item The elapsed time of 12 Myr required for the simulated bubbles
      to reach the desired morphology
      is ten times longer than the typical synchrotron and IC cooling time scale of
      high energy ($\sim100$ GeV) CRe. Thus, re-acceleration of CRe by shocks or turbulence must be considered
      in this model. Since the forward shock is rather far away from the gamma-ray bubbles, stochastic acceleration of CRs by turbulence appears to be more plausible.\
      We will investigate the competition between stochastic acceleration and radiative cooling
      in a future work.
      %{\color{red} KY: This item is to be modified according to the newly added text in the Discussion section. PH: done!}
\end{itemize}

\section{Acknowledgements}
The authors thank Mateusz Ruszkowski and Ellen G. Zweibel for insightful comments.\
We thank Henry Zovaro for discussions about the innermost bubbles\
during the final stages of this work.\
We thank Peter Predehl for providing the range of observed X-ray intensity in Fig. \ref{fig__xray_0.8keV_angle_000}.\
We further thank Chen Chun-Yen provides the code module solving the cosmic ray convection equation.
A part of the simulations are performed and analyzed using computing resources\
operated by the National Center for High-Performance Computing (NCHC).
HYKY acknowledges support from\
Yushan Scholar Program of the Ministry of Education of Taiwan\
and\
Ministry of Science and Technology of Taiwan (MOST 109-2112-M-007-037-MY3).
HS acknowledges\
funding support from the Jade Mountain Young Scholar Award No. NTU-109V0201,\
sponsored by the Ministry of Education, Taiwan.\
This research is partially supported by the Ministry of Science and\
Technology of Taiwan (MOST) under grants MOST 107-2119-M-002-036-MY3\
and MOST 108-2112-M-002-023-MY3, and the NTU\
Core Consortium project under grants NTU-CC-108L893401 and\
NTU-CC-108L893402.

@@

Finally, we find that increasing the source temperature (B$\rightarrow$C$\rightarrow$D)\
not only improves the symmetry of the bubbles but also increases the temperature
inside the bubbles at the present time. This result implies that a high-temperature source
helps the jets ablate the clumpy dense gas away and pours into the atmosphere
through the intercloud channels of the clumpy disk. In contrast,
if source temperature is not high enough,
the jets cannot effectively ablate the high-density clouds within 12 Myr,
resulting in the formation of asymmetric bubbles (case B).
